Certified Symbolic Finite Transducers: Formalization and Applications to String Analysis

### 2025-09-12 Paper/SFT.tex
polish the formalization parts
Working on related work, shortening etc.

\section{Symbolic Finite Transducers}
\label{sec:formalization}

We begin by presenting a mathematical definition of SFTs \cite{VeanesHLMB12Transducer}, abstracting from the specific implementation details of Isabelle/HOL.

@@

SFTs accept an input word and generate an output word. This can be defined by \emph{runs} of SFTs.
An SFT run $\sigma$ is a sequence $(q_0, \phi_0, f_0, q_1),(q_1, \phi_1, f_1, q_2),\ldots, (q_{n-1}, \phi_{n-1}, f_{n-1}, q_n)$ such that $q_0\in \mathcal{I}$ and $(q_i,
\phi_i, f_i, q_{i+1}), 0 \leq i \leq n-1$ is a transition in $\Delta$.
$\sigma$ is an accepting run when $q_n$ is an accepting state.

>>>>

\section{Symbolic Finite Transducers}
\label{sec:sft}

We recall the definition of SFTs \cite{VeanesHLMB12Transducer}, 
abstracting from specific implementation details in Isabelle/HOL.

@@

SFTs accept an input word and generate an output word. This can be defined by \emph{runs} of SFTs.
An SFT run $\sigma$ is a sequence $(q_0, \phi_0, f_0, q_1),(q_1, \phi_1, f_1, q_2),\ldots, (q_{n-1}, \phi_{n-1}, f_{n-1}, q_n)$ such that $q_0\in \mathcal{I}$ and $(q_i,
\phi_i, f_i, q_{i+1}), 0 \leq i \leq n-1$ is a transition in $\Delta$.
$\sigma$ is \emph{accepting} when $q_n \in \mathcal{F}$.

### 2025-09-12 Paper/formalization.tex
Working on related work, shortening etc.

Firstly, $'a \text{ set option }$ is the input type of the transition, it accepts a set of elements of type $'a$ or $\texttt{None}$ corresponding to empty string $\varepsilon$. 
Accepting a set of $'a$ elements aims to express the same but more abstract semantics of the input labels in Definition \ref{def-sft}, in which an input label is a predicate. A predicate's semantics as introduced before represents a set of elements that make the predicate true. But predicates have various different forms. For instance, the interval $[1-9]$ represents the set $\{e \mid 1 \leq e \leq 9\}$. The predicate $\lambda b.~ b[7] = 1$, where $b$ is a bit vector of length 8, denotes the set of bit vectors that have a 1 in the 7th position. All these different forms of labels are abstracted as sets in our formalization.
%
The value $\texttt{None}$ represents the empty string $\varepsilon$ in our formalization, indicating a transition that consumes no input but may still produce output elements. This design choice facilitates the modeling of real-world applications in SFTs, as we will demonstrate in our application to string solvers.

The second element of type $'i$ in LTTS serves as an index into the output function space. 
%
We use indices instead of functions themselves to enable the reuse of the same output functions for different transitions.
%
The mapping $\mathcal{M}_t$ associates each index with a specific output function. These output functions, formalized by \texttt{Tlabel}, map a single input element to a set of possible output elements rather than to a single element. This design enables non-deterministic output behavior, where the transducer may select any element from the output set randomly or according to specified criteria. Additionally, output functions can produce the empty string $\varepsilon$ by returning $\texttt{None}$, providing further flexibility in transition behavior.

>>>>

Firstly, $'a \text{ set option }$ is the input type of the transition, it accepts a set of elements of type $'a$ or $\texttt{None}$ corresponding to empty string $\varepsilon$. 
Accepting a set of $'a$ elements aims to express the same but more abstract semantics of the input labels in Definition \ref{def-sft}, in which an input label is a predicate. A predicate's semantics as introduced before represents a set of elements that make the predicate true. But predicates have various different forms. For instance, the interval $[1-9]$ represents the set $\{e \mid 1 \leq e \leq 9\}$. The predicate $\lambda b.~ b[7] = 1$, where $b$ is a bit vector of length 8, denotes the set of bit vectors that have a 1 in the 7th position. All different forms of labels are abstracted as sets in our formalization.
%
The value $\texttt{None}$ represents the empty string $\varepsilon$ in our formalization, indicating a transition that consumes no input but may still produce output elements. This design choice facilitates the modeling of real-world applications in SFTs, as we will demonstrate in our application to string solvers.

The second element of type $'i$ in LTTS serves as an index into the output function space. 
%
We use indices instead of functions %themselves 
to enable the reuse of the same output functions for different transitions.
%
The mapping $\mathcal{M}_t$ associates each index with a specific output
function. These output functions, formalized by \texttt{Tlabel}, map a single
input element to a set of possible output elements rather than to a single
element. This design enables nondeterministic output behavior, where the
transducer may select any element from the output set arbitrarily or according
to specified criteria. Further, output functions can produce empty string $\varepsilon$ by returning $\texttt{None}$, providing further flexibility in transition behavior.

### 2025-09-12 Paper/introduction.tex
Working on related work, shortening etc.

Owing to the security-critical nature of many applications of string analysis
(e.g. analysis of web applications against XSS vulnerabilities), there is an 
increasing need of trustworthy implementations in such application domains. 
Unfortunately, string analysis implementations are often quite buggy. For
example, even state-of-the-art SMT solvers over the theory of strings 
were found to be buggy through such techniques as fuzzing
\cite{DBLP:conf/cav/BlotskyMBZKG18,BM20,Mansur20}. 
String solvers are among others employed at AWS for analyzing Role-Based 
Access Control (RBAC) policies, which are security-critical.
These have motivated initial formalization efforts for the theory of strings in 
Isabelle/HOL, e.g., a verified solver \cite{cpp/KanLRS22}
for handling a small fragment of SMT-LIB 2.6 unicode theory of strings
consisting only of equality, concatenation and regular constraints, and a
formalization of the semantics of this theory \cite{verified-verifying}.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers (SFTs) in Isabelle/HOL.
We prove the correctness of key operations between SFTs and symbolic SFAs, such as the product operation between SFTs and SFAs.
\item We leverage the refinement framework to develop an extensible and efficient SFT formalization capable of supporting diverse Boolean algebras, ensuring adaptability to future transition label representations. The correctness of the refined, efficient implementation is proved via structural isomorphism between refinement layers.
\item We develop a certified interval algebra with verified set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization by applying it to sanitizer analysis and integrating our formalized SFT into the verified forward-propagation string solver CertiStr \cite{cpp/KanLRS22}. This integration extends CertiStr from supporting only simple concatenation and regular constraints to handling more sophisticated string replacement operations.
\end{enumerate}

>>>>

Owing to the security-critical nature of many applications of string analysis
(e.g. analysis of web applications against XSS vulnerabilities
\cite{Berkeley-JavaScript} and
at AWS for analyzing Role-Based 
Access Control (RBAC) policies \cite{neha,DBLP:conf/fmcad/BackesBCDGLRTV18}), 
there is an 
increasing need of trustworthy implementations in such application domains. 
Unfortunately, string analysis implementations are often quite buggy. For
example, even state-of-the-art SMT solvers over the theory of strings 
were found to be buggy through such techniques as fuzzing
\cite{DBLP:conf/cav/BlotskyMBZKG18,BM20,Mansur20}. 
%Recent applications of string solvers in security critical applications
%(e.g. at AWS for analyzing Role-Based 
%Access Control (RBAC) policies \cite{neha,DBLP:conf/fmcad/BackesBCDGLRTV18})
This
has motivated initial formalization efforts for the theory of strings in 
Isabelle/HOL, e.g., a verified solver \cite{cpp/KanLRS22}
for handling a small fragment of SMT-LIB 2.6 unicode theory of strings
consisting only of equality, concatenation and regular constraints, and a
formalization of the semantics of this theory \cite{verified-verifying}.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers (SFTs) in Isabelle/HOL.
We prove the correctness of key operations of SFTs and symbolic SFAs, such as the product operation between SFTs and SFAs.
\item We leverage the refinement framework to develop an extensible and efficient SFT formalization capable of supporting diverse Boolean algebras, ensuring adaptability to future transition label representations. The correctness of the refined, efficient implementation is proved via structural isomorphism between refinement layers.
\item We develop a certified interval algebra with verified set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization by applying it to sanitizer analysis and integrating our formalized SFT into the verified forward-propagation string solver CertiStr \cite{cpp/KanLRS22}. This integration extends CertiStr from supporting only simple concatenation and regular constraints to handling more sophisticated string replacement operations.
\end{enumerate}

### 2025-09-12 Paper/relatedwork.tex
polish application, conclusion, and related work
Working on related work, shortening etc.

\emph{Symbolic Automata and Transducers.} Symbolic Automata and Transducers \cite{cav/DAntoniV17,VeanesHLMB12Transducer, popl/DAntoniV14, entcs/DAntoniKW18, sofsem/TammV18} represent a significant advancement in automata theory, offering improved efficiency in operations and enhanced expressiveness through algebraic theories that support infinite alphabets. This symbolic framework has been progressively extended to accommodate more complex structures: symbolic tree transducers \cite{ershov/VeanesB11} handle hierarchical data structures, while symbolic pushdown automata \cite{cav/DAntoniA14} manage nested word structures. Recent developments in 2024 have further expanded the scope of symbolic techniques to include B\"{u}chi automata and omega-regular languages \cite{pacmpl/VeanesBEZ25}, enabling verification of infinite-state systems and analysis of non-terminating computations.

\emph{Applications of Symbolic Transducers.} The efficiency, scalability, and expressive power of symbolic automata and transducers have led to their widespread adoption in numerous real-world applications. These applications span diverse domains, including: constraint solving for program analysis \cite{lpar/VeanesBM10}, security-critical sanitizer analysis for web applications \cite{uss/HooimeijerLMSV11}, runtime verification of system behaviors \cite{osdi/YaseenABCL20}, and automated program inversion for software transformation \cite{pldi/HuD17}. Each application leverages the symbolic approach's ability to handle complex patterns and infinite alphabets efficiently.

\emph{Formalization of Symbolic Automata and Transducers.} While classical automata theories have been extensively formalized in interactive theorem provers \cite{Tuerk-NFA, Lammich2014TheCA, Peter14, cpp/DoczkalKS13}, with some work on transducer formalization \cite{afp/LochmannFSTS21}, the symbolic variants of automata and transducers remain largely unexplored in formal verification. To our knowledge, CertiStr \cite{cpp/KanLRS22} represents the only existing work on symbolic automata formalization in a proof assistant. Our work advances this frontier by extending the formal treatment to both SFTs and $\varepsilon$SFAs.

\emph{String Solving.} A significant application of our certified transducer framework is in string constraint solving, a field that has seen intensive research development over the past decade. While our work provides formal verification guarantees, there exists a rich ecosystem of non-certified string solvers, each with distinct capabilities: Kaluza \cite{Berkeley-JavaScript} specializes in JavaScript analysis, CVC5 \cite{cvc5}, Z3-str3 \cite{Z3-str3} builds on the Z3 framework, S3P \cite{DBLP:conf/ccs/TrinhCJ14}, Ostrich \cite{pacmpl/ChenFHHHKLRW22}, and SLOTH \cite{HJLRV18}.
As more and more bugs have beein uncovered in existing string solvers \cite{DBLP:conf/cav/BlotskyMBZKG18} and SMT solvers \cite{Mansur20} 
We believe that our work will benefit the community by providing a formal foundation for string solvers development.

\emph{Certified SMT Solvers.} Beyond string theories, certification efforts in SMT solving have extended to other domains. For example, the work by Shi et al. \cite{DBLP:conf/cav/ShiFLTWY20}, who developed a certified SMT solver for quantifier-free bit-vector theory, demonstrating the broader applicability of interactive theorem proving in certified SMT solver development.

>>>>

\emph{Symbolic Automata and Transducers.} Symbolic Automata and Transducers \cite{cav/DAntoniV17,VeanesHLMB12Transducer, popl/DAntoniV14, entcs/DAntoniKW18, sofsem/TammV18} represent a significant advancement in automata theory, offering improved efficiency in operations and enhanced expressiveness through algebraic theories that support infinite alphabets. This symbolic framework has been progressively extended to accommodate more complex structures: symbolic tree transducers \cite{ershov/VeanesB11} handle hierarchical data structures, while symbolic pushdown automata \cite{cav/DAntoniA14} manage nested word structures. 
%Recent developments in 2024 have further expanded the scope of symbolic techniques to include B\"{u}chi automata and omega-regular languages \cite{pacmpl/VeanesBEZ25}, enabling verification of infinite-state systems and analysis of non-terminating computations.

\emph{Applications of Symbolic Transducers.} The efficiency, scalability, and
expressive power of symbolic automata and transducers have led to their
widespread adoption in numerous applications, including: constraint solving for program analysis \cite{lpar/VeanesBM10}, security-critical sanitizer analysis for web applications \cite{uss/HooimeijerLMSV11}, runtime verification of system behaviors \cite{osdi/YaseenABCL20}, automated program inversion for software 
transformation \cite{pldi/HuD17}, and string solving
\cite{pacmpl/ChenFHHHKLRW22,CHL+19}. Each application leverages the symbolic 
approach's ability to handle complex
patterns and large alphabets efficiently.

\emph{Formalization of Symbolic Automata and Transducers.} While classical automata theories have been extensively formalized in interactive theorem provers \cite{Tuerk-NFA, Lammich2014TheCA, Peter14, cpp/DoczkalKS13}, with some work on transducer formalization \cite{afp/LochmannFSTS21}, the symbolic variants of automata and transducers remain largely unexplored in formal verification. 
%To our knowledge, CertiStr \cite{cpp/KanLRS22} represents the only existing work on symbolic automata formalization in a proof assistant. Our work advances this frontier by extending the formal treatment to both SFTs and $\varepsilon$SFAs.

\emph{String Solving.} A significant application of our certified transducer
framework is in string constraint solving, a field that has seen intensive
research development over the past decade. While our work provides formal
verification guarantees, there exists a rich ecosystem of non-certified string
solvers, which has in the last few years led to a standardized SMT-LIB 2.6/2.7 
unicode string theory and annual competition. SMT-LIB compatible string 
solvers include Z3 \cite{z3}, Z3-alpha \cite{z3alpha}, Z3-Noodler 
\cite{noodler,noodler-int,noodler-len, noodler-tool}, Z3str3RE \cite{z3str3re},
cvc5 \cite{cvc5}, Z3str4 \cite{z3str4}, Trau \cite{trau0,trau-tool,trau2}, and 
OSTRICH \cite{ostrich}. Recent SMT competition over string theory has revealed
superiority of automata-based methods (based on restrictions of symbolic
automata). For this reason,
%: Kaluza \cite{Berkeley-JavaScript} specializes in JavaScript analysis, CVC5 \cite{cvc5}, Z3-str3 \cite{Z3-str3} builds on the Z3 framework, S3P \cite{DBLP:conf/ccs/TrinhCJ14}, Ostrich \cite{pacmpl/ChenFHHHKLRW22}, and SLOTH \cite{HJLRV18}.
%As more and more 
%Numerous bugs have been uncovered in existing string solvers \cite{DBLP:conf/cav/BlotskyMBZKG18} and SMT solvers \cite{Mansur20} 
our work will benefit the community by providing a formal 
foundation for string solvers development.

\emph{Certified SMT Solvers.} Beyond string theories, certification efforts in SMT solving have extended to other domains. For example, the work by Shi et al. \cite{DBLP:conf/cav/ShiFLTWY20}, who developed a certified SMT solver for quantifier-free bit-vector theory, demonstrating the broader applicability of interactive theorem proving in certified SMT solver development. The work by Mansur et al. \cite{verified-verifying} define a formal semantics for SMT-LIB string
to check compatibility between solvers and SMTLIB standard.
